Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-20 introduction.tex
lloid_paper: trim and rework the EM followup section anticipating that table 1 will merge into Figure 2
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1124 ccdd05d5-56c1-461a-bf98-85f6876523b4

\begin{table}[h]
\caption{\label{table:sky-localization-accuracy}Horizon distance, expected detection rate, and \SNR\ at merger for sources that produce an \SNR\ of 8 in all three detectors at time $t$ before merger.  The fifth and sixth columns show the minimum area of the 90\% confidence region before merger and at merger.}
\begin{center}
\begin{tabular}{rrrrrr}
\tableline\tableline
& horiz. & rate & final & \multicolumn{2}{c}{$A$(90\%) (deg$^2$)} \\
\cline{5-6}
$t$ (s) & (Mpc) & (yr$^{-1}$) & \SNR\ & early & final \\
\tableline
25.0 & 226 & 5 & 15.8 & 2893.2 & 1.8 \\
10.0 & 299 & 12 & 11.9 & 1150.6 & 3.6 \\
5.0 & 341 & 18 & 10.4 & 594.6 & 5.1 \\
1.0 & 401 & 29 & 8.9 & 153.9 & 7.5 \\
0.1 & 435 & 37 & 8.2 & 32.5 & 9.1 \\
0.0 & 445 & 40 & 8.0 & 9.6 & 9.6 \\
\tableline
\end{tabular}
\end{center}
\end{table}
%
\begin{figure}[h]
\plotone{figures/loc_in_time.pdf}
\caption{}
\end{figure}
%
\EM{} followup requires estimating the location of the \GW{} source. The
localization uncertainty is dominated by uncertainty in the time of arrival of
the \GW{}s, which is determined by the signal's effective bandwidth and \SNR{}
\citep{Fairhurst2009}.  Table~\ref{table:sky-localization-accuracy} shows the
minimum 90\% confidence area for a source that produces an \SNR\ of 8 in the
\LIGO\ detectors in Hanford, Washington, and Livingston, Louisiana, and, the
Virgo detector in Cascina, Italy, at a time $t$ before merger.  Also shown are
the final \SNR\ and 90\% confidence area at merger for the same source.
At 10~s before merger, the pre-merger 90\% confidence area is about $1150.6$~deg$^2$
(2.8\% of the sky) and at 1~s before merger is 153.9~deg$^2$ (0.37\% of the sky).
The confidence region shrinks quadratically in area as the overall amplitude of the
signal increases.

It is infeasible to search this large an area for a prompt counterpart. For
reference, \textsc{lsst} should see first light around the same time as
Advanced \LIGO{} and will have an unparalleled field of view of 9.6 deg$^2$
\citep{2008arXiv0805.2366I}.  However, if we can take galactic luminosities
to accurately reflect mass and thus be a proxy for the \CBC{} event rate, it is
possible to reduce the localization uncertainty using a galaxy catalog, along
with the sky location and luminosity distance estimate from the \GW{} signal,
as a prior probability distribution \citep{galaxy-catalog} as was done in
S6/VSR3. Within the expected Advanced \LIGO{} \textsc{ns}--\textsc{ns} horizon
distance, the number of galaxies that can produce a given signal
amplitude is much larger than the number for initial \LIGO{} and thus the
catalog will not be as useful for downselecting pointings generally. However,
sources that are detectable before merger are likely to be exceptionally
nearby. Thus there will be fewer galaxies to consider for a given
early-warning candidate and catalog completeness will be less of a concern.
This should allow us to shrink the 90\% confidence area substantially.
Even so, it will be desirable to mobilize a large number of heterogeneous
telescopes in order to guarantee imaging this exceptional event at the instant
of merger. Because the sky localization will improve rapidly until merger,
by the time that the entire \CBC{} event is over it will be
possible to send a second alert to downselect the images that have already been
taken as well as repoint telescopes to more likely locations.

>>>>

\begin{table}[h]
\caption{\label{table:sky-localization-accuracy}Horizon distance, expected detection rate, and \SNR\ at merger for sources that produce an \SNR\ of 8 in all three detectors at time $t$ before merger.  The fifth and sixth columns show the minimum area of the 90\% confidence region before merger and at merger.}
\begin{center}
\begin{tabular}{rrrrrr}
\tableline\tableline
& horiz. & rate & final & \multicolumn{2}{c}{$A$(90\%) (deg$^2$)} \\
\cline{5-6}
$t$ (s) & (Mpc) & (yr$^{-1}$) & \SNR\ & early & final \\
\tableline
25.0 & 226 & 5 & 15.8 & 2893.2 & 1.8 \\
10.0 & 299 & 12 & 11.9 & 1150.6 & 3.6 \\
5.0 & 341 & 18 & 10.4 & 594.6 & 5.1 \\
1.0 & 401 & 29 & 8.9 & 153.9 & 7.5 \\
0.1 & 435 & 37 & 8.2 & 32.5 & 9.1 \\
0.0 & 445 & 40 & 8.0 & 9.6 & 9.6 \\
\tableline
\end{tabular}
\end{center}
\end{table}
%
\begin{figure}[h]
\plotone{figures/loc_in_time.pdf}
\caption{\label{fig:sky-localization-accuracy} Area of the 90\% confidence
region as a function of time for an exceptional source, where such a source is
defined to produce a signal with an \SNR\ of at least 27.4. These signals are
expected to be observed in the detectors once per year.  80~s, 35~s, 16~s, and
6~s before coalescence, such a signal will have accumulated an \SNR\ of 8, 12,
16, and 20, respectively. After merger, the minimum 90\% confidence area
shrinks to $0.82$~deg$^2$, shown by the dashed line.}
\end{figure}

\EM{} followup requires estimating the location of the \GW{} source. The
localization uncertainty is dominated by uncertainty in the time of arrival of
the \GW{}s, which is determined by the signal's effective bandwidth and \SNR{}
\citep{Fairhurst2009}. Figure~\ref{fig:sky-localization-accuracy} estimates the
minimum 90\% confidence area versus time for a few exceptionally loud
coalescence events detectable by Advanced \LIGO{} and Advanced Virgo. A
once-per-year event would have minimum 90\% confidence areas of $2250$~deg$^2$
($X$\% of the sky) and $270$~deg$^2$ ($Y$\% of the sky) at times 30~s and 10~s
before coalescence, respectively. The confidence region shrinks quadratically in
area as the overall amplitude of the signal increases. By the time of merger,
the minimum 90\% confidence area shrinks to $0.82$~deg$^2$.

It is infeasible to search this hundreds of square degrees for a prompt
counterpart. For
reference, \textsc{lsst} should see first light around the same time as
Advanced \LIGO{} and will have an unparalleled field of view of 9.6 deg$^2$
\citep{2008arXiv0805.2366I}.  However, if we can take galactic luminosities
to accurately reflect mass and thus be a proxy for the \CBC{} event rate, it is
possible to reduce the localization uncertainty using a galaxy catalog, along
with the sky location and luminosity distance estimate from the \GW{} signal,
as a prior probability distribution \citep{galaxy-catalog} as was done in
S6/VSR3. Within the expected Advanced \LIGO{} \textsc{ns}--\textsc{ns} horizon
distance, the number of galaxies that can produce a given signal
amplitude is much larger than the number for initial \LIGO{} and thus the
catalog will not be as useful for downselecting pointings generally. However,
our exceptional sources will necessarily be extremely nearby. Within this
reduced volume there will be fewer galaxies to consider for a given
candidate and catalog completeness will be less of a concern.
This should allow us to shrink the 90\% confidence area substantially.
